---
title: "Accessing cMD data releases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accessing cMD data releases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# The code needs the live site, so it is shown, not run, when the package builds.
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This is a walk through the public curatedMetagenomicData (cMD) releases for
the cMD team: find a release, list its studies, get a species
`TreeSummarizedExperiment` for a study, query whole tables with SQL or dplyr,
and download HUMAnN gene families.

```{r}
library(cmgdr)
```

## Datasets and releases

Each pipeline registration (a pipeline version plus its configuration) is its
own **dataset**, named `<workflow_id>-<version>`:

| dataset | what it holds |
|---|---|
| `cmgd_nextflow-2.2.1` | pipeline 2.2.1: MetaPhlAn 4.2.2, Bracken, resistome, markers, QC; samples keyed by the md5 `sample_id` |
| `cmgd_mpa4.2-2.3.0` | pipeline 2.3.0 without HUMAnN; samples keyed by readset id (`RS.…`) |
| `cmgd_humann3.9-2.3.0` | as above, plus HUMAnN 3.9 pathways and gene families |
| `cmgd_humann4a1-2.3.0` | as above with HUMAnN 4.0.0a1 |

A dataset is published as **releases**: immutable full snapshots named by
their build date. Cite the dataset and the release id.

```{r}
cmgd_datasets()
cmgd_releases("cmgd_nextflow-2.2.1")
```

Open a release. `"latest"` (the default) resolves to the newest one; pin a
release id for reproducible analyses.

```{r}
rel <- cmgd_release("cmgd_nextflow-2.2.1")
rel <- cmgd_release("cmgd_nextflow-2.2.1", release = "2026-10-08")
cmgd_tables(rel)
```

## Studies

Each release has ready-made files per study: a species × samples matrix, the
study's rows of the main tables (Parquet) and a QC table.

```{r}
studies <- cmgd_studies(rel)
head(studies[order(-studies$n_samples), ])
cmgd_study_files(rel, "ZellerG_2014")
```

`cmgd_read_study()` downloads one (into a verified cache, see below) and reads
it. Parquet files need the duckdb package.

```{r}
qc <- cmgd_read_study(rel, "ZellerG_2014", "qc")
species <- cmgd_read_study(rel, "ZellerG_2014", "metaphlan_species")
bracken <- cmgd_read_study(rel, "ZellerG_2014", "bracken")
```

## A species TreeSummarizedExperiment

`cmgd_species_tse()` turns a study's species matrix into a sparse
`TreeSummarizedExperiment`: relative abundance in percent from the main
MetaPhlAn pass on all reads, one column per sample in `qc.tsv`, with the QC
table as `colData`.

```{r}
tse <- cmgd_species_tse(rel, "ZellerG_2014")
tse
SummarizedExperiment::colData(tse)$reads_decontaminated
```

## Whole tables with SQL and dplyr

`cmgd_connect()` attaches the release's read-only DuckLake catalog in DuckDB.
Nothing is downloaded up front: DuckDB fetches only the parts of the Parquet
files a query touches.

```{r}
con <- cmgd_connect(rel)
DBI::dbGetQuery(con, "
  SELECT study_name, count(*) AS n_samples, avg(reads_decontaminated) AS mean_reads
  FROM qc_metrics
  GROUP BY study_name
  ORDER BY n_samples DESC")
```

The same with dplyr. A species missing from a sample has no row, so divide by
the study's sample count rather than taking a mean.

```{r}
library(dplyr)
n <- tbl(con, "qc_metrics") |> filter(study_name == "ZellerG_2014") |> count() |> pull(n)
tbl(con, "taxonomic_profile_metaphlan") |>
  filter(study_name == "ZellerG_2014", rank == "species",
         data_type == "full_data", is.na(humann_bundle)) |>
  group_by(clade_name) |>
  summarise(mean_percent = sum(relative_abundance) / n) |>
  arrange(desc(mean_percent)) |>
  head(10) |>
  collect()
DBI::dbDisconnect(con, shutdown = TRUE)
```

Before computing across tables:

1. **Abundances are not comparable across methods.** MetaPhlAn
   `relative_abundance` is a percent; Bracken `fraction_total_reads` is a 0–1
   fraction.
2. **Pick one `data_type`**: `full_data` (all reads) or `rarefied_data` (a
   1M-read subsample). Most analyses want `full_data`.
3. **Pick one MetaPhlAn pass.** In HUMAnN datasets the bundle's own MetaPhlAn
   pass (`humann_bundle` set) sits next to the main pass
   (`humann_bundle IS NULL`).

Each release's `README.md` and `tables/<table>/schema.json` document every
column.

## Files and the cache

`cmgd_files()` lists the Parquet files behind each table, and
`cmgd_download()` fetches any listed file (or the DuckLake catalog) by its
release-relative path or URL. Downloads land in `cmgd_cache_dir()`, are
checked against the release's SHA-256 checksums (a mismatch is an error and the
file is discarded), and are not fetched again while the cached copy verifies.

```{r}
files <- cmgd_files(rel, "qc_metrics")
files
local <- cmgd_download(rel, files$path[1])
```

## HUMAnN gene families

Gene families are too large for the tables, so each sample's HUMAnN
gene-family table is its own download from `https://cmgd-raw.cancerdatasci.org`,
listed in the HUMAnN datasets' gene-family index. Values are HUMAnN's
unnormalized output in the bundle's units.

```{r}
gf_rel <- cmgd_release("cmgd_humann3.9-2.3.0")
cmgd_genefamilies(gf_rel)  # studies, sample counts
gf <- cmgd_genefamilies(gf_rel, study = "ZellerG_2014")
gf[order(-gf$size), c("sample_key", "size", "rows")]
paths <- vapply(gf$url[1:2], function(u) cmgd_download(gf_rel, u), character(1))
genefamilies <- read.delim(paths[1], check.names = FALSE)
```

The release lists gene families per study, so asking for one study fetches
only that study's index. Downloads are checked against the index's size and
SHA-256 (computed when the file was ingested).
